---
title: "Epigenica vs Illumina Using CpG Islands"
author: "Zoe dell'Osso"
date: "March 2026"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE)
```

---

# 1. Extracting CpG island data for Epigenica data

```{bash}
# Copying bed file with all CpG islands to working directory
cp ~/EpivsIllumina_analysis/CpG_island.bed .

# Copying bigwig files from D1 R1 to working directory
cp ~/Epigenica_data/Zoe_analysis/analysis_bam/*Donor1_S1* .

# Copying bigwig files from D1 R2 to working directory
cp ~/Epigenica_data/Zoe_analysis/analysis_bam/*Donor1_S2* .

# Extracting CpG island data from Epigenica bigwig files — D1 R1
multiBigwigSummary BED-file \
  --bwfiles M2_S2_Donor1_S1_pooled.hg19.bw \
            M3_S3_Donor1_S1_pooled.hg19.bw \
            M4_S4_Donor1_S1_pooled.hg19.bw \
  --BED CpG_island.bed \
  --labels MBDIH Cxxc MBDCO \
  --outRawCounts Epi_raw_CpGi.tab \
  -o results.npz

# Extracting CpG island data from Epigenica bigwig files — D1 R2
multiBigwigSummary BED-file \
  --bwfiles M2_S2_Donor1_S2_pooled.hg19.bw \
            M3_S3_Donor1_S2_pooled.hg19.bw \
            M4_S4_Donor1_S2_pooled.hg19.bw \
  --BED CpG_island.bed \
  --labels MBDIH Cxxc MBDCO \
  --outRawCounts Epi_raw_CpGi_D1R2.tab \
  -o resultsD1R2.npz

# Checking how many rows are in the tab files
wc -l Epi_raw_CpGi.tab
# 27842
wc -l Epi_raw_CpGi_D1R2.tab
# 27842
```

```{r import-epi-data}
# Importing the tab files into R
D1_R1_Epi_data <- read.table("Epi_raw_CpGi.tab",    header = TRUE)
D1_R2_Epi_data <- read.table("Epi_raw_CpGiD1R2.tab", header = TRUE)

# Viewing how the data looks before renaming
head(D1_R1_Epi_data)

# Adding column names
colnames(D1_R1_Epi_data) <- c("chr", "Start", "End", "MBDIH",    "CXXC",    "MBDCO")
colnames(D1_R2_Epi_data) <- c("chr", "Start", "End", "MBDIH_R2", "CXXC_R2", "MBDCO_R2")

# Viewing how the data looks after renaming
head(D1_R1_Epi_data)
head(D1_R2_Epi_data)

# Checking the number of rows
nrow(D1_R1_Epi_data)
# [1] 27840
nrow(D1_R2_Epi_data)
# [1] 27840
```

**Output — head(D1_R1_Epi_data) before renaming:**

```
  chr  Start    End MBDIH CXXC MBDCO
1 chr1 135124 135563     0    0     0
2 chr1 327790 328229     0    0     0
3 chr1 437151 438164     0    0     0
4 chr1 449273 450544     0    0     0
5 chr1 533219 534114     0    0     0
6 chr1 544738 546649     0    0     0
```

**Output — head(D1_R2_Epi_data) after renaming:**

```
   chr  Start    End MBDIH_R2 CXXC_R2 MBDCO_R2
1 chr1 135124 135563        0       0        0
2 chr1 327790 328229        0       0        0
3 chr1 437151 438164        0       0        0
4 chr1 449273 450544        0       0        0
5 chr1 533219 534114        0       0        0
6 chr1 544738 546649        0       0        0
```

---

# 2. Calculating beta values for both MBDs

```{r calculate-beta}
# Load library
library(dplyr)

# Beta values for in house
D1_R1_Epi_data$Beta_valueIH    <- D1_R1_Epi_data$MBDIH    / (D1_R1_Epi_data$MBDIH    + D1_R1_Epi_data$CXXC)
D1_R2_Epi_data$Beta_valueIH_R2 <- D1_R2_Epi_data$MBDIH_R2 / (D1_R2_Epi_data$MBDIH_R2 + D1_R2_Epi_data$CXXC_R2)

# Beta values for commercial
D1_R1_Epi_data$Beta_valueCO    <- D1_R1_Epi_data$MBDCO    / (D1_R1_Epi_data$MBDCO    + D1_R1_Epi_data$CXXC)
D1_R2_Epi_data$Beta_valueCO_R2 <- D1_R2_Epi_data$MBDCO_R2 / (D1_R2_Epi_data$MBDCO_R2 + D1_R2_Epi_data$CXXC_R2)

# Testing how many beta values are NaN (no reads captured by both CXXC and MBD)
# In house
sum(is.nan(D1_R1_Epi_data$Beta_valueIH))
# [1] 954
sum(is.nan(D1_R2_Epi_data$Beta_valueIH_R2))
# [1] 956

# Commercial
sum(is.nan(D1_R1_Epi_data$Beta_valueCO))
# [1] 955
sum(is.nan(D1_R2_Epi_data$Beta_valueCO_R2))
# [1] 958
```

Since the number of CpG islands with NaN values differs between the in-house and commercial MBD, two separate data frames must be created to ensure correct exclusion.

```{r remove-nan-beta}
# Removing CpG islands with NaN beta values

# In house
D1_R1_EpiIH_data <- D1_R1_Epi_data %>%
  filter(!is.na(Beta_valueIH)) %>%
  select(-Beta_valueCO, -MBDCO)

D1_R2_EpiIH_data <- D1_R2_Epi_data %>%
  filter(!is.na(Beta_valueIH_R2)) %>%
  select(-Beta_valueCO_R2, -MBDCO_R2)

# Checking if it worked
sum(is.nan(D1_R1_EpiIH_data$Beta_valueIH))
# [1] 0
nrow(D1_R1_EpiIH_data)
# [1] 26886
nrow(D1_R2_EpiIH_data)
# [1] 26884

# Commercial
D1_R1_EpiCO_data <- D1_R1_Epi_data %>%
  filter(!is.na(Beta_valueCO)) %>%
  select(-Beta_valueIH, -MBDIH)

D1_R2_EpiCO_data <- D1_R2_Epi_data %>%
  filter(!is.na(Beta_valueCO_R2)) %>%
  select(-Beta_valueIH_R2, -MBDIH_R2)

# Checking if it worked
sum(is.nan(D1_R1_EpiCO_data$Beta_valueCO))
# [1] 0
nrow(D1_R1_EpiCO_data)
# [1] 26885
nrow(D1_R2_EpiCO_data)
# [1] 26882
```

The number of rows decreases in line with the NaN values that were identified and removed.

---

# 3. Identifying CpG islands to exclude from Epigenica data

## 3.1 Filtering out unknown chromosome locations

```{r filter-unknown-chr}
# Load library
library(dplyr)

# Removing CpG islands on unknown chromosome locations

# In house
D1_R1_EpiIH_data <- D1_R1_EpiIH_data %>% filter(!grepl("chrUn", chr))
D1_R2_EpiIH_data <- D1_R2_EpiIH_data %>% filter(!grepl("chrUn", chr))

# Checking if it worked
nrow(D1_R1_EpiIH_data)
# [1] 26790
nrow(D1_R2_EpiIH_data)
# [1] 26788

# Commercial
D1_R1_EpiCO_data <- D1_R1_EpiCO_data %>% filter(!grepl("chrUn", chr))
D1_R2_EpiCO_data <- D1_R2_EpiCO_data %>% filter(!grepl("chrUn", chr))

# Checking if it worked
nrow(D1_R1_EpiCO_data)
# [1] 26789
nrow(D1_R2_EpiCO_data)
# [1] 26786
```

A total of 96 CpG islands were removed from both the in-house and commercial data frames.

## 3.2 Histogram of read count vs number of CpG islands

```{r readcount-histogram}
# Load libraries
library(ggplot2)
library(dplyr)

# For in house — calculating the sum of read counts (Epigenica IH and CXXC)
D1_R1_EpiIH_data$sum_mbdih_n_cxxc <- D1_R1_EpiIH_data$MBDIH + D1_R1_EpiIH_data$CXXC

# Generating histogram with filtered sum, 30 binwidth
readcounts_distribution <- ggplot(D1_R1_EpiIH_data, aes(x = sum_mbdih_n_cxxc)) +
  geom_histogram(fill = "red", color = "white", binwidth = 30) +
  labs(title = "Distribution of Combined Read Counts",
       x     = "Total read counts (MBD in house and CXXC)",
       y     = "Number of Islands (Frequency)") +
  theme_minimal()

# Saving the histogram as a PNG
ggsave("Distribution_readcounts_mbhih_n_cxxc.png",
       plot = readcounts_distribution, width = 10, height = 6, dpi = 300)

# For commercial — calculating the sum of read counts (Epigenica CO and CXXC)
D1_R1_EpiCO_data$sum_mbdco_n_cxxc <- D1_R1_EpiCO_data$MBDCO + D1_R1_EpiCO_data$CXXC

# Generating histogram with filtered sum, 30 binwidth
readcounts_distribution_CO <- ggplot(D1_R1_EpiCO_data, aes(x = sum_mbdco_n_cxxc)) +
  geom_histogram(fill = "red", color = "white", binwidth = 30) +
  labs(title = "Distribution of Combined Read Counts",
       x     = "Total read counts (MBD commercial and CXXC)",
       y     = "Number of Islands (Frequency)") +
  theme_minimal()

# Saving the histogram as a PNG
ggsave("Distribution_readcounts_mbhco_n_cxxc.png",
       plot = readcounts_distribution_CO, width = 10, height = 6, dpi = 300)
```

---

# 4. Normalization of Epigenica beta values

```{r normalize-epi-beta}
# Load libraries
library(limma)
library(ggplot2)
library(dplyr)

# In house — merging the replicates
D1_EpiIH_data <- merge(D1_R1_EpiIH_data, D1_R2_EpiIH_data, by = c("chr", "Start", "End"))

# Ensuring beta values are numeric (columns 6 and 11)
matrix_D1_EpiIH_data <- as.matrix(D1_EpiIH_data)
cols_to_norm          <- matrix_D1_EpiIH_data[, c(6, 11)]
numeric_cols          <- apply(cols_to_norm, 2, as.numeric)

# Normalizing using cyclic loess
norm_D1_EpiIH_data <- normalizeBetweenArrays(numeric_cols, method = "cyclicloess")

# Putting the normalized beta values back into the data frame
D1_EpiIH_data[, c(6, 11)] <- norm_D1_EpiIH_data

# Commercial — merging the replicates
D1_EpiCO_data <- merge(D1_R1_EpiCO_data, D1_R2_EpiCO_data, by = c("chr", "Start", "End"))

# Ensuring beta values are numeric (columns 6 and 11)
matrix_D1_EpiCO_data <- as.matrix(D1_EpiCO_data)
cols_to_norm_CO       <- matrix_D1_EpiCO_data[, c(6, 11)]
numeric_cols_CO       <- apply(cols_to_norm_CO, 2, as.numeric)

# Normalizing using cyclic loess
norm_D1_EpiCO_data <- normalizeBetweenArrays(numeric_cols_CO, method = "cyclicloess")

# Putting the normalized beta values back into the data frame
D1_EpiCO_data[, c(6, 11)] <- norm_D1_EpiCO_data
```

## 4.1 Overlaying histograms — read counts vs beta values

### In House

```{r overlay-hist-ih}
# Load libraries
library(ggplot2)
library(tidyr)

# Converting df to long format for the histogram
longdf_D1_EpiIH_data <- D1_EpiIH_data %>%
  pivot_longer(cols = c(MBDIH, CXXC), names_to = "Method", values_to = "Read_Counts")

# Creating histogram
histogram_cxxc_mbdih <- ggplot(longdf_D1_EpiIH_data,
                                aes(x = Beta_valueIH, weight = Read_Counts, fill = Method)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 50, color = "white") +
  scale_fill_manual(values = c("MBDIH" = "#E41A1C", "CXXC" = "#377EB8")) +
  labs(title = "Overlaying Histogram: Total Read Counts vs Beta Values",
       x     = "Beta Value",
       y     = "Total Read Counts (MBD in house + CXXC)") +
  theme_minimal()

# Saving the histogram as a PNG
ggsave("Histogram_D1_R1_readcounts_MBDIH_n_CXXC_vs_bvalue.png",
       plot = histogram_cxxc_mbdih, width = 10, height = 6, dpi = 300)
```

### Commercial

```{r overlay-hist-co}
# Load libraries
library(ggplot2)
library(tidyr)

# Converting df to long format for the histogram
longdf_D1_EpiCO_data <- D1_EpiCO_data %>%
  pivot_longer(cols = c(MBDCO, CXXC), names_to = "Method", values_to = "Read_Counts")

# Creating histogram
histogram_cxxc_mbdco <- ggplot(longdf_D1_EpiCO_data,
                                aes(x = Beta_valueCO, weight = Read_Counts, fill = Method)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 50, color = "white") +
  scale_fill_manual(values = c("MBDCO" = "#E41A1C", "CXXC" = "#377EB8")) +
  labs(title = "Overlaying Histogram: Total Read Counts vs Beta Values",
       x     = "Beta Value",
       y     = "Total Read Counts (MBD commercial + CXXC)") +
  theme_minimal()

# Saving the histogram as a PNG
ggsave("Histogram_D1_R1_readcounts_MBDCO_n_CXXC_vs_bvalue.png",
       plot = histogram_cxxc_mbdco, width = 10, height = 6, dpi = 300)
```

---

# 5. Illumina Array data processing

## 5.1 Normalizing Illumina values with NOOB

```{r noob-normalization}
# Load library
library(minfi)

# Creating a path to the folder that has the Illumina data
input_dir <- "/home/ubuntu/Illumina_850k_DNAmethylation_data_donors/data/"

# Creating a data frame from sample information
targets <- read.metharray.sheet(input_dir)
head(targets)

# Loading raw methylation data into R
rgSet <- read.metharray.exp(base = input_dir)

# Normalizing the data with Noob
mSet <- preprocessNoob(rgSet)

# Creating matrix with beta values for CpG regions
noob_beta <- getBeta(mSet)
head(noob_beta)

# Changing column names
colnames(noob_beta) <- targets$Sample_Name
head(noob_beta)

# Saving beta values to move to CpG islands analysis directory
write.csv(noob_beta, "Final_NOOB_Normalized_Beta_Matrix.csv")
```

```{bash}
cp ~/Illumina_850k_DNAmethylation_data_donors/Zoe_analysis/Final_NOOB_Normalized_Beta_Matrix.csv .
```

```{r import-noob}
# Importing the NOOB normalized beta values
noob_beta <- read.csv("Final_NOOB_Normalized_Beta_Matrix.csv")
```

**Output — head(targets):**

```
  Sample_Name Sample_Group Sex Age_predicted  Array        Slide
1          D1        Donor   M            54 R01C01 205061410164
2          D2        Donor   M            52 R02C01 205061410164
3          D3        Donor   M            62 R03C01 205061410164
4          D4        Donor   M            47 R04C01 205061410164
```

**Output — head(noob_beta) after renaming columns:**

```
           D1         D2         D3         D4
cg18478105 0.01489857 0.01550465 0.01977528 0.01706442
cg09835024 0.03834064 0.04149480 0.04331674 0.03732362
cg14361672 0.91584946 0.92746948 0.92231030 0.91727131
cg01763666 0.72587764 0.73984052 0.78786225 0.82832947
cg12950382 0.94589055 0.94160644 0.94003364 0.93680127
cg02115394 0.06638772 0.07629724 0.10196685 0.10088353
```

## 5.2 Getting the Illumina beta values for each CpG island

```{r illumina-cpg-islands}
# Load libraries
library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
library(dplyr)

# Extracting CpG island data with Illumina annotations
data(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
data(Islands.UCSC)
force(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
force(Islands.UCSC)
island_df <- as.data.frame(Islands.UCSC)
head(island_df)

# Making Illumina probes the row names
rownames(noob_beta) <- noob_beta$X

# Merging the Illumina beta data with the island data by row names
noob_ill_islands <- merge(island_df, noob_beta, by = 0)
head(noob_ill_islands)

# Checking number of rows
nrow(noob_ill_islands)
# [1] 865859

# Filtering to keep only probes within a CpG island
noob_ill_islands <- noob_ill_islands %>%
  filter(Relation_to_Island == "Island")
head(noob_ill_islands)

# Checking number of rows after filtering
nrow(noob_ill_islands)
# [1] 161441

# Extracting only Donor 1
D1_noob_islands <- noob_ill_islands[, c("Row.names", "Islands_Name", "D1")]

# Grouping rows by Island_Name
D1_noob_islands <- D1_noob_islands %>% group_by(Islands_Name)

# Checking number of different CpG islands captured
n_groups(D1_noob_islands)
# [1] 26594

# Averaging the D1 values for each group
D1_noob_islands_averages <- D1_noob_islands %>%
  summarise(Average_D1 = mean(D1, na.rm = TRUE))

# Checking number of rows
nrow(D1_noob_islands_averages)
# [1] 26594
```

**Output — head(noob_ill_islands) after island filtering:**

```
  Row.names              Islands_Name Relation_to_Island         D1         D2         D3         D4
1 cg00000622   chr15:23033933-23034710             Island 0.02359467 0.02395984 0.02536818 0.02846882
2 cg00000658  chr9:139997846-139998048             Island 0.80039323 0.77342384 0.79437498 0.68826105
3 cg00000734  chr3:128902134-128902978             Island 0.04318268 0.06751227 0.07456778 0.06830980
4 cg00000769 chr12:124086217-124087015             Island 0.01214208 0.01403837 0.01402864 0.01658373
5 cg00000807    chr2:23913308-23913569             Island 0.75326514 0.77810563 0.75014132 0.80778165
6 cg00000924     chr11:2720410-2722087             Island 0.52557253 0.48536452 0.52401268 0.47780166
```

**Output — head(D1_noob_islands_averages):**

```
# A tibble: 6 × 2
  Islands_Name                  Average_D1
  <chr>                              <dbl>
1 chr10:100028204-100028508         0.0885
2 chr10:100227438-100227832         0.0379
3 chr10:100992156-100992687         0.129
4 chr10:100993820-100994188         0.0172
5 chr10:101089010-101090655         0.369
6 chr10:101190451-101190925         0.0659
```

---

# 6. Overlaying Histograms — Beta Values for Epigenica and Illumina

## 6.1 Merging beta value data into 1 data frame

```{r merge-beta-values}
# Load library
library(dplyr)

# Creating Islands_Name column from chromosome coordinates in Epi data sets
D1_EpiIH_data <- D1_EpiIH_data %>%
  mutate(Islands_Name = paste0(chr, ":", Start, "-", End))

D1_EpiCO_data <- D1_EpiCO_data %>%
  mutate(Islands_Name = paste0(chr, ":", Start, "-", End))

# Joining data sets using inner_join — only Island_Names present in all are kept
D1_R1_merged_data <- D1_EpiIH_data %>%
  inner_join(D1_EpiCO_data,            by = "Islands_Name") %>%
  inner_join(D1_noob_islands_averages, by = "Islands_Name") %>%
  select(Islands_Name, Average_D1, Beta_valueIH, Beta_valueCO)
```

## 6.2 In House vs Illumina

```{r hist-epiIH-illumina}
# Load libraries
library(ggplot2)
library(tidyr)

# Creating a long df for the histogram
longdf_D1_R1_merged_cpgi_epiIH <- pivot_longer(
  D1_R1_merged_data,
  cols      = c("Average_D1", "Beta_valueIH"),
  names_to  = "Variable",
  values_to = "Value")

# Plotting histogram
histogram_D1_R1_merged_cpgi_epiIH <- ggplot(
  longdf_D1_R1_merged_cpgi_epiIH,
  aes(x = Value, fill = Variable)) +
  geom_histogram(bins = 30, color = "white", alpha = 0.5, position = "identity") +
  scale_fill_manual(
    values = c("Average_D1"   = "#F8766D", "Beta_valueIH" = "#619CFF"),
    labels = c("Average_D1"   = "Illumina",  "Beta_valueIH" = "Epigenica IH")) +
  labs(title = "Distribution of Beta Values for Illumina and Epigenica IH",
       x     = "Beta Values",
       y     = "Frequency (Number of CpG Islands)",
       fill  = "Technologies") +
  theme_minimal()

# Saving as PNG
ggsave("Histogram_D1_R1_epiIH_Ill_bvalues_CpGislands_noob.png",
       plot = histogram_D1_R1_merged_cpgi_epiIH, width = 10, height = 6, dpi = 300)
```

## 6.3 Commercial vs Illumina

```{r hist-epiCO-illumina}
# Load libraries
library(ggplot2)
library(tidyr)

# Creating a long df for the histogram
longdf_D1_R1_merged_cpgi_epiCO <- pivot_longer(
  D1_R1_merged_data,
  cols      = c("Average_D1", "Beta_valueCO"),
  names_to  = "Variable",
  values_to = "Value")

# Plotting histogram
histogram_D1_R1_merged_cpgi_epiCO <- ggplot(
  longdf_D1_R1_merged_cpgi_epiCO,
  aes(x = Value, fill = Variable)) +
  geom_histogram(bins = 30, color = "white", alpha = 0.5, position = "identity") +
  scale_fill_manual(
    values = c("Average_D1"  = "#F8766D", "Beta_valueCO" = "#619CFF"),
    labels = c("Average_D1"  = "Illumina",  "Beta_valueCO" = "Epigenica CO")) +
  labs(title = "Distribution of Beta Values for Illumina and Epigenica CO",
       x     = "Beta Values",
       y     = "Frequency (Number of CpG Islands)",
       fill  = "Technologies") +
  theme_minimal()

# Saving as PNG
ggsave("Histogram_D1_R1_epiCO_Ill_bvalues_CpGislands_noob.png",
       plot = histogram_D1_R1_merged_cpgi_epiCO, width = 10, height = 6, dpi = 300)
```

## 6.4 In House vs Commercial

```{r hist-epiIH-epiCO}
# Load libraries
library(ggplot2)
library(tidyr)

# Creating a long df for the histogram
longdf_D1_R1_merged_cpgi_epiCO_and_epiIH <- pivot_longer(
  D1_R1_merged_data,
  cols      = c("Beta_valueIH", "Beta_valueCO"),
  names_to  = "Variable",
  values_to = "Value")

# Plotting histogram
histogram_D1_R1_merged_cpgi_epiCO_and_epiIH <- ggplot(
  longdf_D1_R1_merged_cpgi_epiCO_and_epiIH,
  aes(x = Value, fill = Variable)) +
  geom_histogram(bins = 30, color = "white", alpha = 0.5, position = "identity") +
  scale_fill_manual(
    values = c("Beta_valueIH" = "#F8766D", "Beta_valueCO" = "#619CFF"),
    labels = c("Beta_valueIH" = "Epigenica IH", "Beta_valueCO" = "Epigenica CO")) +
  labs(title = "Distribution of Beta Values for Epigenica IH and CO",
       x     = "Beta Values",
       y     = "Frequency (Number of CpG Islands)",
       fill  = "Technologies") +
  theme_minimal()

# Saving as PNG
ggsave("Histogram_D1_R1_epiIH_epiCO_bvalues_CpGislands.png",
       plot = histogram_D1_R1_merged_cpgi_epiCO_and_epiIH, width = 10, height = 6, dpi = 300)
```

---

# 7. Confusion Matrix

## 7.1 Three Classes

### In House

```{r cm-3class-ih}
# Load libraries
library(dplyr)
library(tidyr)
library(caret)
library(ggplot2)

# Creating the 3 classes with 0.33 intervals
my_breaks2 <- c(0, 0.33, 0.66, 1.0)
my_labels2 <- c("Class 1", "Class 2", "Class 3")

# Adding classes to the beta values
matrixdf_class3_D1_R1_epiIH_ill <- D1_R1_merged_data %>%
  mutate(
    Epi_Class     = cut(Beta_valueIH, breaks = my_breaks2, labels = my_labels2, include.lowest = TRUE),
    Illumina_Class = cut(Average_D1,   breaks = my_breaks2, labels = my_labels2, include.lowest = TRUE))

# Visualizing the classes assigned to the beta values
head(matrixdf_class3_D1_R1_epiIH_ill)

# Making confusion matrix
conf_matrix_class3_D1_R1_epiIH_ill <- confusionMatrix(
  data      = matrixdf_class3_D1_R1_epiIH_ill$Epi_Class,
  reference = matrixdf_class3_D1_R1_epiIH_ill$Illumina_Class)

print(conf_matrix_class3_D1_R1_epiIH_ill)

# Plotting the confusion matrix
plotdata_cm_class3_D1_R1_epiIH_ill <- as.data.frame(conf_matrix_class3_D1_R1_epiIH_ill$table)

plot_cm_class3_D1_R1_epiIH_ill <- ggplot(
  plotdata_cm_class3_D1_R1_epiIH_ill,
  aes(x = Prediction, y = Reference, fill = Freq)) +
  geom_tile(color = "white", lwd = 1) +
  geom_text(aes(label = Freq), color = "black", size = 5) +
  scale_fill_gradient(low = "#9ecae1", high = "#084594") +
  labs(x     = "Epi IH",
       y     = "Illumina",
       title = "Confusion Matrix: Epi IH vs Illumina",
       fill  = "Count") +
  theme_minimal() +
  theme(axis.text = element_text(size = 12, face = "bold"),
        title     = element_text(size = 14, face = "bold"))

# Saving as a PNG
ggsave("D1_R1_confusion_matrix_c3_epiIH_ill_noob.png",
       plot = plot_cm_class3_D1_R1_epiIH_ill, width = 7, height = 6, dpi = 300)
```

**Output — head(matrixdf_class3_D1_R1_epiIH_ill):**

```
              Islands_Name Average_D1 Beta_valueIH Beta_valueCO Epi_Class Illumina_Class
1   chr1:10003165-10003585 0.08212839  0.018950395  0.012946372   Class 1        Class 1
2     chr1:1002663-1005318 0.41891865  0.530955968  0.148345027   Class 2        Class 2
3 chr1:100315420-100316009 0.05619820  0.009532714  0.006069872   Class 1        Class 1
4 chr1:100435297-100436070 0.10577649  0.013251458  0.010354383   Class 1        Class 1
5 chr1:100503482-100504404 0.04831715  0.015095061  0.002821707   Class 1        Class 1
6   chr1:10057121-10058108 0.13466606  0.035888133  0.018354921   Class 1        Class 1
```

**Output — Confusion matrix (3 classes, In House):**

```
Confusion Matrix and Statistics

          Reference
Prediction Class 1 Class 2 Class 3
   Class 1   16085    1056      14
   Class 2     107     576     106
   Class 3      11     772    6199

Overall Statistics
               Accuracy : 0.9171
                 95% CI : (0.9136, 0.9205)
    No Information Rate : 0.65
    P-Value [Acc > NIR] : < 2.2e-16
                  Kappa : 0.8268
 Mcnemar's Test P-Value : < 2.2e-16

Statistics by Class:
                     Class: Class 1 Class: Class 2 Class: Class 3
Sensitivity                  0.9927        0.23960         0.9810
Specificity                  0.8773        0.99054         0.9579
Pos Pred Value               0.9376        0.73004         0.8879
Neg Pred Value               0.9848        0.92427         0.9933
Prevalence                   0.6500        0.09645         0.2535
Detection Rate               0.6453        0.02311         0.2487
Detection Prevalence         0.6882        0.03165         0.2801
Balanced Accuracy            0.9350        0.61507         0.9695
```

### Commercial

```{r cm-3class-co}
# Load libraries
library(dplyr)
library(tidyr)
library(caret)
library(ggplot2)

# Creating the 3 classes with 0.33 intervals
my_breaks2 <- c(0, 0.33, 0.66, 1.0)
my_labels2 <- c("Class 1", "Class 2", "Class 3")

# Adding classes to the beta values
matrixdf_class3_D1_R1_epiCO_ill <- D1_R1_merged_data %>%
  mutate(
    Epi_Class      = cut(Beta_valueCO, breaks = my_breaks2, labels = my_labels2, include.lowest = TRUE),
    Illumina_Class = cut(Average_D1,   breaks = my_breaks2, labels = my_labels2, include.lowest = TRUE))

# Visualizing the classes assigned to the beta values
head(matrixdf_class3_D1_R1_epiCO_ill)

# Making confusion matrix
conf_matrix_class3_D1_R1_epiCO_ill <- confusionMatrix(
  data      = matrixdf_class3_D1_R1_epiCO_ill$Epi_Class,
  reference = matrixdf_class3_D1_R1_epiCO_ill$Illumina_Class)

print(conf_matrix_class3_D1_R1_epiCO_ill)

# Plotting the confusion matrix
plotdata_cm_class3_D1_R1_epiCO_ill <- as.data.frame(conf_matrix_class3_D1_R1_epiCO_ill$table)

plot_cm_class3_D1_R1_epiCO_ill <- ggplot(
  plotdata_cm_class3_D1_R1_epiCO_ill,
  aes(x = Prediction, y = Reference, fill = Freq)) +
  geom_tile(color = "white", lwd = 1) +
  geom_text(aes(label = Freq), color = "black", size = 5) +
  scale_fill_gradient(low = "#9ecae1", high = "#084594") +
  labs(x     = "Epi CO",
       y     = "Illumina",
       title = "Confusion Matrix: Epi CO vs Illumina",
       fill  = "Count") +
  theme_minimal() +
  theme(axis.text = element_text(size = 12, face = "bold"),
        title     = element_text(size = 14, face = "bold"))

# Saving as a PNG
ggsave("D1_R1_confusion_matrix_c3_epiCO_ill_noob.png",
       plot = plot_cm_class3_D1_R1_epiCO_ill, width = 7, height = 6, dpi = 300)
```

**Output — Confusion matrix (3 classes, Commercial):**

```
Confusion Matrix and Statistics

          Reference
Prediction Class 1 Class 2 Class 3
   Class 1   15760    1363     114
   Class 2       8     318     443
   Class 3       2     163    6156

Overall Statistics
               Accuracy : 0.914
                 95% CI : (0.9104, 0.9175)
    No Information Rate : 0.6483
    P-Value [Acc > NIR] : < 2.2e-16
                  Kappa : 0.8156
 Mcnemar's Test P-Value : < 2.2e-16

Statistics by Class:
                     Class: Class 1 Class: Class 2 Class: Class 3
Sensitivity                  0.9994        0.17245         0.9170
Specificity                  0.8274        0.97994         0.9906
Pos Pred Value               0.9143        0.41352         0.9739
Neg Pred Value               0.9986        0.93522         0.9691
Prevalence                   0.6483        0.07580         0.2759
Detection Rate               0.6478        0.01307         0.2531
Detection Prevalence         0.7086        0.03161         0.2598
Balanced Accuracy            0.9134        0.57620         0.9538
```

## 7.2 Ten Classes

### In House

```{r cm-10class-ih}
# Load libraries
library(dplyr)
library(tidyr)
library(caret)
library(ggplot2)

# Creating 10 classes with 0.1 intervals
my_breaks3 <- c(0, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1.0)
my_labels3 <- c("Class 1", "Class 2", "Class 3", "Class 4", "Class 5",
                 "Class 6", "Class 7", "Class 8", "Class 9", "Class 10")

# Classifying the beta values into 10 classes
matrixdf_class10_D1_R1_epiIH_ill <- D1_R1_merged_data %>%
  mutate(
    Epi_Class      = cut(Beta_valueIH, breaks = my_breaks3, labels = my_labels3, include.lowest = TRUE),
    Illumina_Class = cut(Average_D1,   breaks = my_breaks3, labels = my_labels3, include.lowest = TRUE))

# Making confusion matrix
conf_matrix_class10_D1_R1_epiIH_ill <- confusionMatrix(
  data      = matrixdf_class10_D1_R1_epiIH_ill$Epi_Class,
  reference = matrixdf_class10_D1_R1_epiIH_ill$Illumina_Class)

# Converting the confusion matrix table into a df
plotdata_cm_class10_D1_R1_epiIH_ill <- as.data.frame(conf_matrix_class10_D1_R1_epiIH_ill$table)

# Plotting the confusion matrix
plot_cm_class10_D1_R1_epiIH_ill <- ggplot(
  plotdata_cm_class10_D1_R1_epiIH_ill,
  aes(x = Prediction, y = Reference, fill = Freq)) +
  geom_tile(color = "white", lwd = 1) +
  geom_text(aes(label = Freq), color = "black", size = 3.5) +
  scale_fill_gradient(low = "#ebf3fb", high = "#3182bd") +
  labs(x     = "Epi IH",
       y     = "Illumina",
       title = "Confusion Matrix: Epi IH vs Illumina",
       fill  = "Count") +
  theme_minimal() +
  theme(axis.text    = element_text(angle = 45, vjust = 1, hjust = 1),
        title        = element_text(size = 14, face = "bold"),
        aspect.ratio = 1)

# Saving as a PNG
ggsave("D1_R1_confusion_matrix_c10_epiIH_ill_noob.png",
       plot = plot_cm_class10_D1_R1_epiIH_ill, width = 7, height = 6, dpi = 300)
```

### Commercial

```{r cm-10class-co}
# Load libraries
library(dplyr)
library(tidyr)
library(caret)
library(ggplot2)

# Creating 10 classes with 0.1 intervals
my_breaks3 <- c(0, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1.0)
my_labels3 <- c("Class 1", "Class 2", "Class 3", "Class 4", "Class 5",
                 "Class 6", "Class 7", "Class 8", "Class 9", "Class 10")

# Classifying the beta values into 10 classes
matrixdf_class10_D1_R1_epiCO_ill <- D1_R1_merged_data %>%
  mutate(
    Epi_Class      = cut(Beta_valueCO, breaks = my_breaks3, labels = my_labels3, include.lowest = TRUE),
    Illumina_Class = cut(Average_D1,   breaks = my_breaks3, labels = my_labels3, include.lowest = TRUE))

# Making confusion matrix
conf_matrix_class10_D1_R1_epiCO_ill <- confusionMatrix(
  data      = matrixdf_class10_D1_R1_epiCO_ill$Epi_Class,
  reference = matrixdf_class10_D1_R1_epiCO_ill$Illumina_Class)

# Converting the confusion matrix table into a df
plotdata_cm_class10_D1_R1_epiCO_ill <- as.data.frame(conf_matrix_class10_D1_R1_epiCO_ill$table)

# Plotting the confusion matrix
plot_cm_class10_D1_R1_epiCO_ill <- ggplot(
  plotdata_cm_class10_D1_R1_epiCO_ill,
  aes(x = Prediction, y = Reference, fill = Freq)) +
  geom_tile(color = "white", lwd = 1) +
  geom_text(aes(label = Freq), color = "black", size = 3.5) +
  scale_fill_gradient(low = "#ebf3fb", high = "#3182bd") +
  labs(x     = "Epi CO",
       y     = "Illumina",
       title = "Confusion Matrix: Epi CO vs Illumina",
       fill  = "Count") +
  theme_minimal() +
  theme(axis.text    = element_text(angle = 45, vjust = 1, hjust = 1),
        title        = element_text(size = 14, face = "bold"),
        aspect.ratio = 1)

# Saving as a PNG
ggsave("D1_R1_confusion_matrix_c10_epiCO_ill_noob.png",
       plot = plot_cm_class10_D1_R1_epiCO_ill, width = 7, height = 6, dpi = 300)
```
