---
title: "CpG Sites Analysis - Comparing EpiFinder to Illumina MethylationEPIC 850K Array"
author: "Zoe dell'Osso"
date: "March 2026"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE)
```

---

# 1. QC and Processing

## 1.1 QC Epigenica data using BAM files

```{bash}
# To get a fingerprint with input, MBD in house, CXXC, MBD commercial for D1 S1
plotFingerprint \
  -b *Donor1_S1*.bam \
  --labels INPUT MBD_in_house CXXC MBD_commercial \
  --minMappingQuality 30 \
  --skipZeros \
  -T "QC: Donor 1 sample 1" \
  -o fingerprint_qc.png

# To generate heatmap using all samples
multiBamSummary bins \
    --bamfiles *.bam \
    -o readCounts.npz \
    --outRawCounts readCounts.txt \
    --numberOfProcessors max

# To generate spearman correlation using all samples
plotCorrelation \
  -in readCounts.npz \
  --corMethod spearman \
  --whatToPlot heatmap \
  --plotTitle "Spearman Correlation of Samples" \
  -o heatmap_correlation.png \
  --colorMap RdYlBu

# To generate the PCA plot using all samples
plotPCA \
    -in readCounts.npz \
    -o plot_PCA.png \
    --plotTitle "PCA Plot"
```

## 1.2 QC EPIC Array data using beta values with Quantile normalization

```{r array-qc}
# Illumina EPIC array data processing

# Load library
library(minfi)

# Creating a path to the folder that has the Illumina data
input_dir <- "/home/ubuntu/Illumina_850k_DNAmethylation_data_donors/data/"

# Creating a data frame from sample information
targets <- read.metharray.sheet(input_dir)

# Load raw methylation data into R
rgSet <- read.metharray.exp(base = input_dir)

# Normalize the data with Quantile
mSet_Qunorm <- preprocessQuantile(rgSet)

# Create matrix with beta values for CpG sites
beta_norQuat <- getBeta(mSet_Qunorm)

# Change column names
colnames(beta_norQuat) <- targets$Sample_Name

# View name of columns and data
head(beta_norQuat)

# Changing to a data frame
beta_norQuat <- as.data.frame(beta_norQuat)

# Adding ProbeIDs as a column
beta_norQuat$ProbeIDs <- rownames(beta_norQuat)
# 865859 rows

# Generating QC plots

# Plot density of Beta-values and save as a PNG
png("Beta_Density_Plot.png", width = 800, height = 600)
densityPlot(beta_norQuat, main = "Beta Value Density Plot")
dev.off()

# Creating a MDS plot and saving as a PNG
png("Beta_MDS_Plot.png", width = 800, height = 600)
mdsPlot(beta_norQuat, numPositions = 1000, sampGroups = NULL, main = "MDS Plot (Beta Matrix)")
dev.off()
```

**Output — head(beta_norQuat):**

```
                  D1        D2        D3        D4
cg14817997 0.7716823 0.7553782 0.7180083 0.6851830
cg26928153 0.7736478 0.7795466 0.8260223 0.7816682
cg16269199 0.6781921 0.6415911 0.6735047 0.5992393
cg13869341 0.7041774 0.7547598 0.7692744 0.7448073
cg14008030 0.5933294 0.6081829 0.6667299 0.6078450
cg12045430 0.2765107 0.2849464 0.2787795 0.2900669
```

## 1.3 Beta values calculations for EPIC Array without normalization

```{r beta-no-norm}
# Load library
library(minfi)

# Beta values without normalization
beta_NOnorm <- getBeta(rgSet)
colnames(beta_NOnorm) <- targets$Sample_Name
beta_NOnorm <- as.data.frame(beta_NOnorm)
beta_NOnorm$ProbeIDs <- rownames(beta_NOnorm)

# Identifying number of rows
nrow(beta_NOnorm)
# 866238
```

## 1.4 Epigenica data processing

### Extracting reads for each EPIC Array 850K probe

```{bash}
# Extraction for M2
multiBigwigSummary BED-file -b M2*.bw --BED EPIC.hg19.manifest3forIGV.bed \
  -out results_D1_S1.npz --outRawCounts mbd_ih_with_probes.tab

# Extraction for M3
multiBigwigSummary BED-file -b M3*.bw --BED EPIC.hg19.manifest3forIGV.bed \
  -out results_cxxc_all.npz --outRawCounts Cxxc_with_probes.tab

# Extraction for M4
multiBigwigSummary BED-file -b M4*.bw --BED EPIC.hg19.manifest3forIGV.bed \
  -out results_M4_all.npz --outRawCounts mdb_co_with_probes.tab

# All had the same bin number of 865906

# Merging M2 & M3
cut -f4-16 Cxxc_with_probes.tab | paste mbd_ih_with_probes.tab - > mbdih_cxxc_probes.tsv

# Merging M3 & M4
cut -f4-16 Cxxc_with_probes.tab | paste mdb_co_with_probes.tab - > mbdco_cxxc_probes.tsv
```

### Inputting Epigenica processed data into R

```{r input-epigenica}
# Creating column names
Columnnames_mbd <- c("chr", "start", "end",
                     "MBD_D1_R1", "MBD_D1_R2",
                     "MBD_D2_R1", "MBD_D2_R2",
                     "MBD_D3_R1", "MBD_D3_R2",
                     "MBD_D4_R1", "MBD_D4_R2",
                     "Cxxc_D1_R1", "Cxxc_D1_R2",
                     "Cxxc_D2_R1", "Cxxc_D2_R2",
                     "Cxxc_D3_R1", "Cxxc_D3_R2",
                     "Cxxc_D4_R1", "Cxxc_D4_R2")

# Inputting the TSV as a data frame for in house
MBDIH_Cxxc <- read.table("mbdih_cxxc_probes.tsv", sep = "\t",
                          header = FALSE, col.names = Columnnames_mbd)

# Inputting the TSV as a data frame for commercial
MBDCO_Cxxc <- read.table("mbdco_cxxc_probes.tsv", sep = "\t",
                          header = FALSE, col.names = Columnnames_mbd)

# Changing chromosome location to Illumina probes and adding ProbeIDs
columnnames_bedfile <- c("chr", "start", "end", "ProbeIDs")
Probes_location <- read.table("EPIC.hg19.manifest3forIGV.bed", sep = "\t",
                               header = FALSE, col.names = columnnames_bedfile)

# 1. MBD in house & CXXC
# Merge data frames based on the three location columns
MBDIH_Cxxc_with_IDs <- merge(Probes_location, MBDIH_Cxxc, by = c("chr", "start", "end"))
# Remove the coordinate columns
MBDIH_final <- MBDIH_Cxxc_with_IDs[, !(names(MBDIH_Cxxc_with_IDs) %in% c("chr", "start", "end"))]
# number of rows = 865906

# 2. MBD commercial & CXXC
# Merge data frames based on the three location columns
MBDCO_Cxxc_with_IDs <- merge(Probes_location, MBDCO_Cxxc, by = c("chr", "start", "end"))
# Remove the coordinate columns
MBDCO_final <- MBDCO_Cxxc_with_IDs[, !(names(MBDCO_Cxxc_with_IDs) %in% c("chr", "start", "end"))]
# number of rows = 865906
```

---

# 2. M value analysis

## 2.1 Histogram — M values without any filtering or normalization

```{r m-values-no-norm}
# Load libraries
library(ggplot2)
library(tidyr)
library(dplyr)
library(DMRcate)
library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)

# Merging the D1 values from Epi IH and array D1
M_value_df_NOnorm <- MBDIH_final[, c("ProbeIDs", "MBD_D1_R1", "Cxxc_D1_R1")]
M_value_df_NOnorm <- M_value_df_NOnorm %>%
  left_join(beta_NOnorm %>% select(ProbeIDs, D1), by = "ProbeIDs")

# Calculating the M value for array
M_value_df_NOnorm$M_values_Ill <- log2(M_value_df_NOnorm$D1 / (1 - M_value_df_NOnorm$D1))

# Calculating the M value for Epigenica data MBD IH
lib_ratio <- sum(M_value_df_NOnorm$MBD_D1_R1) / sum(M_value_df_NOnorm$Cxxc_D1_R1)
M_value_df_NOnorm$M_value_Epi <- log2(
  (M_value_df_NOnorm$MBD_D1_R1 + 1) /
  ((M_value_df_NOnorm$Cxxc_D1_R1 * lib_ratio) + 1))

# View results
head(M_value_df_NOnorm)

# Set a seed so random results are reproducible
set.seed(67)

# Creating df for 150 random probes
M_value_df_NOnorm_150probes <- M_value_df_NOnorm %>%
  slice_sample(n = 150) %>%
  select(M_values_Ill, M_value_Epi) %>%
  mutate(RowNumber = row_number()) %>%
  pivot_longer(cols = c(M_values_Ill, M_value_Epi),
               names_to  = "Source",
               values_to = "Intensity") %>%
  mutate(Source = recode(Source,
                         "M_values_Ill" = "Ill",
                         "M_value_Epi"  = "Epi IH"))

# Plotting the histogram
hist_M_values_NOnorm <- ggplot(M_value_df_NOnorm_150probes,
                                aes(x = RowNumber, y = Intensity, fill = Source)) +
  geom_col(position = "identity", alpha = 0.6) +
  scale_fill_manual(values = c("Ill" = "#F8766D", "Epi IH" = "#619CFF"),
                    name = "Technology") +
  theme_minimal() +
  labs(title = "M-Values for 150 Probes",
       x = "CpG Site",
       y = "M-Value Intensity")

ggsave("Histogram_Mvalues_NOnorm_150probes.png",
       plot = hist_M_values_NOnorm, width = 10, height = 6, dpi = 300)
```

**Output — head(M_value_df_NOnorm):**

```
    ProbeIDs MBD_D1_R1 Cxxc_D1_R1        D1 M_values_Ill M_value_Epi
1 cg19251656  0.000000          1 0.5859689    0.5010852  -0.5580252
2 cg06383512  2.000000          1 0.6634173    0.9789555   1.0269373
3 cg14071171  7.000000          1 0.7418642    1.5230246   2.4419748
4 cg23599391  1.666667          0 0.7384615    1.4974997   1.4150375
5 cg10841437  0.000000          1 0.3293142   -1.0261721  -0.5580252
6 cg06838840  0.000000          2 0.7875618    1.8903498  -0.9594031
```

## 2.2 Spearman Ranked correlation — M values without filtering or normalization

```{r spearman-no-norm}
correlation_Mval_NOnorm <- cor(M_value_df_NOnorm$M_values_Ill,
                                M_value_df_NOnorm$M_value_Epi,
                                method = "spearman",
                                use    = "complete.obs")
print(correlation_Mval_NOnorm)
# 0.5988374
```

## 2.3 Histogram — M values with Quantile normalization

```{r m-values-qunorm}
# Merging the D1 values from Epi IH and array D1
M_value_df_Qunorm <- MBDIH_final[, c("ProbeIDs", "MBD_D1_R1", "Cxxc_D1_R1")]
M_value_df_Qunorm <- M_value_df_Qunorm %>%
  left_join(beta_norQuat %>% select(ProbeIDs, D1), by = "ProbeIDs")

# Calculating the M value for array
M_value_df_Qunorm$M_values_Ill <- log2(M_value_df_Qunorm$D1 / (1 - M_value_df_Qunorm$D1))

# Calculating the M value for Epigenica data MBD IH
lib_ratio <- sum(M_value_df_Qunorm$MBD_D1_R1) / sum(M_value_df_Qunorm$Cxxc_D1_R1)
M_value_df_Qunorm$M_value_Epi <- log2(
  (M_value_df_Qunorm$MBD_D1_R1 + 1) /
  ((M_value_df_Qunorm$Cxxc_D1_R1 * lib_ratio) + 1))

# View results
head(M_value_df_Qunorm)

# Set a seed so random results are reproducible
set.seed(67)

# Creating df for 150 random probes
M_value_df_Qunorm_150probes <- M_value_df_Qunorm %>%
  slice_sample(n = 150) %>%
  select(M_values_Ill, M_value_Epi) %>%
  mutate(RowNumber = row_number()) %>%
  pivot_longer(cols = c(M_values_Ill, M_value_Epi),
               names_to  = "Source",
               values_to = "Intensity") %>%
  mutate(Source = recode(Source,
                         "M_values_Ill" = "Ill",
                         "M_value_Epi"  = "Epi IH"))

# Plotting the histogram
hist_M_values_Qunorm <- ggplot(M_value_df_Qunorm_150probes,
                                aes(x = RowNumber, y = Intensity, fill = Source)) +
  geom_col(position = "identity", alpha = 0.6) +
  scale_fill_manual(values = c("Ill" = "#F8766D", "Epi IH" = "#619CFF"),
                    name = "Technology") +
  theme_minimal() +
  labs(title = "M-Values for 150 Probes",
       x = "CpG Site",
       y = "M-Value Intensity")

ggsave("Histogram_Mvalues_Qunorm_150probes.png",
       plot = hist_M_values_Qunorm, width = 10, height = 6, dpi = 300)
```

**Output — head(M_value_df_Qunorm):**

```
    ProbeIDs MBD_D1_R1 Cxxc_D1_R1        D1 M_values_Ill M_value_Epi
1 cg19251656  0.000000          1 0.5938438    0.5480491  -0.5580252
2 cg06383512  2.000000          1 0.6787592    1.0792449   1.0269373
3 cg14071171  7.000000          1 0.7575530    1.6436776   2.4419748
4 cg23599391  1.666667          0 0.7513807    1.5956057   1.4150375
5 cg10841437  0.000000          1 0.3233789   -1.0651224  -0.5580252
6 cg06838840  0.000000          2 0.7995882    1.9962900  -0.9594031
```

## 2.4 Spearman Ranked correlation — M Values with Quantile Normalization

```{r spearman-qunorm}
correlation_Mval_Qunorm <- cor(M_value_df_Qunorm$M_values_Ill,
                                M_value_df_Qunorm$M_value_Epi,
                                method = "spearman",
                                use    = "complete.obs")
print(correlation_Mval_Qunorm)
# 0.5798922
```

## 2.5 Bland–Altman plot with M values

Lin's Concordance Correlation Coefficient (CCC) measures how well two variables agree with each other.

```{r bland-altman}
# Calculating Lin's concordance correlation coefficient (CCC)
calculate_ccc <- function(x, y) {
  mu_x <- mean(x, na.rm = TRUE)
  mu_y <- mean(y, na.rm = TRUE)
  s2_x <- var(x, na.rm = TRUE)
  s2_y <- var(y, na.rm = TRUE)
  s_xy <- cov(x, y, use = "complete.obs")
  ccc  <- (2 * s_xy) / (s2_x + s2_y + (mu_x - mu_y)^2)
  return(ccc)
}

ccc_val   <- calculate_ccc(M_value_df_Qunorm$M_values_Ill, M_value_df_Qunorm$M_value_Epi)
ccc_label <- paste("CCC =", round(ccc_val, 4))
print(ccc_label)
# CCC = 0.6397 — as the value is <0.9, the strength of agreement is weak.

# Subsample of 5K probes (800K is too many for a standard plot)
set.seed(123)
idx    <- sample(1:nrow(M_value_df_Qunorm), 5000)
df_sub <- M_value_df_Qunorm[idx, ]

# Plotting Bland-Altman plot
png("M_values_Ill_vs_M_values_Epi_Agreement.png", width = 800, height = 800, res = 120)

plot(df_sub$M_values_Ill, df_sub$M_value_Epi,
     main = "Agreement: Ill vs Epi using M values (Subsampled n=5000)",
     xlab = "Ill Intensity",
     ylab = "Epi Intensity",
     pch  = 16,
     col  = rgb(0.2, 0.4, 0.6, 0.2),
     xlim = c(-8, 8),
     ylim = c(-8, 8))

# Adding line of agreement and legend with CCC value
abline(0, 1, col = "red", lwd = 2, lty = 2)
legend("topleft", legend = ccc_label, bty = "n", cex = 1.2, text.font = 2)
dev.off()
```

---

# 3. Beta values analysis

## 3.1 Calculating beta values for all donors and replicates

### In House

```{r beta-inhouse}
# Donor 1 Replicate 1
MBDIH_final$Beta_D1_R1 <- MBDIH_final$MBD_D1_R1 / (MBDIH_final$MBD_D1_R1 + MBDIH_final$Cxxc_D1_R1)

# Donor 1 Replicate 2
MBDIH_final$Beta_D1_R2 <- MBDIH_final$MBD_D1_R2 / (MBDIH_final$MBD_D1_R2 + MBDIH_final$Cxxc_D1_R2)

# Donor 2 Replicate 1
MBDIH_final$Beta_D2_R1 <- MBDIH_final$MBD_D2_R1 / (MBDIH_final$MBD_D2_R1 + MBDIH_final$Cxxc_D2_R1)

# Donor 2 Replicate 2
MBDIH_final$Beta_D2_R2 <- MBDIH_final$MBD_D2_R2 / (MBDIH_final$MBD_D2_R2 + MBDIH_final$Cxxc_D2_R2)

# Donor 3 Replicate 1
MBDIH_final$Beta_D3_R1 <- MBDIH_final$MBD_D3_R1 / (MBDIH_final$MBD_D3_R1 + MBDIH_final$Cxxc_D3_R1)

# Donor 3 Replicate 2
MBDIH_final$Beta_D3_R2 <- MBDIH_final$MBD_D3_R2 / (MBDIH_final$MBD_D3_R2 + MBDIH_final$Cxxc_D3_R2)

# Donor 4 Replicate 1
MBDIH_final$Beta_D4_R1 <- MBDIH_final$MBD_D4_R1 / (MBDIH_final$MBD_D4_R1 + MBDIH_final$Cxxc_D4_R1)

# Donor 4 Replicate 2
MBDIH_final$Beta_D4_R2 <- MBDIH_final$MBD_D4_R2 / (MBDIH_final$MBD_D4_R2 + MBDIH_final$Cxxc_D4_R2)

# Copying Beta columns into a new data frame
Betavalues_MBDIH <- MBDIH_final[, c("ProbeIDs",
                                     "Beta_D1_R1", "Beta_D1_R2",
                                     "Beta_D2_R1", "Beta_D2_R2",
                                     "Beta_D3_R1", "Beta_D3_R2",
                                     "Beta_D4_R1", "Beta_D4_R2")]
nrow(Betavalues_MBDIH)
# 865906 rows

# Delete rows with NaN (no reads captured by either MBD or CXXC)
Betavalues_MBDIH <- na.omit(Betavalues_MBDIH)
# After deleting: 264914 rows left
```

### Commercial

```{r beta-commercial}
# Donor 1 Replicate 1
MBDCO_final$Beta_D1_R1 <- MBDCO_final$MBD_D1_R1 / (MBDCO_final$MBD_D1_R1 + MBDCO_final$Cxxc_D1_R1)

# Donor 1 Replicate 2
MBDCO_final$Beta_D1_R2 <- MBDCO_final$MBD_D1_R2 / (MBDCO_final$MBD_D1_R2 + MBDCO_final$Cxxc_D1_R2)

# Donor 2 Replicate 1
MBDCO_final$Beta_D2_R1 <- MBDCO_final$MBD_D2_R1 / (MBDCO_final$MBD_D2_R1 + MBDCO_final$Cxxc_D2_R1)

# Donor 2 Replicate 2
MBDCO_final$Beta_D2_R2 <- MBDCO_final$MBD_D2_R2 / (MBDCO_final$MBD_D2_R2 + MBDCO_final$Cxxc_D2_R2)

# Donor 3 Replicate 1
MBDCO_final$Beta_D3_R1 <- MBDCO_final$MBD_D3_R1 / (MBDCO_final$MBD_D3_R1 + MBDCO_final$Cxxc_D3_R1)

# Donor 3 Replicate 2
MBDCO_final$Beta_D3_R2 <- MBDCO_final$MBD_D3_R2 / (MBDCO_final$MBD_D3_R2 + MBDCO_final$Cxxc_D3_R2)

# Donor 4 Replicate 1
MBDCO_final$Beta_D4_R1 <- MBDCO_final$MBD_D4_R1 / (MBDCO_final$MBD_D4_R1 + MBDCO_final$Cxxc_D4_R1)

# Donor 4 Replicate 2
MBDCO_final$Beta_D4_R2 <- MBDCO_final$MBD_D4_R2 / (MBDCO_final$MBD_D4_R2 + MBDCO_final$Cxxc_D4_R2)

# Copying Beta columns into a new data frame
Betavalues_MBDCO <- MBDCO_final[, c("ProbeIDs",
                                     "Beta_D1_R1", "Beta_D1_R2",
                                     "Beta_D2_R1", "Beta_D2_R2",
                                     "Beta_D3_R1", "Beta_D3_R2",
                                     "Beta_D4_R1", "Beta_D4_R2")]
nrow(Betavalues_MBDCO)
# 865906 rows

# Delete rows with NaN (no reads captured by either MBD or CXXC)
Betavalues_MBDCO <- na.omit(Betavalues_MBDCO)
# After deleting: 366135 rows left
```

---

# 4. Comparison using ATB samples

## 4.1 Calculating delta beta

```{r delta-beta}
# Calculate average beta value per donor — MBD In House
Betavalues_MBDIH$Average_D1 <- (Betavalues_MBDIH$Beta_D1_R1 + Betavalues_MBDIH$Beta_D1_R2) / 2
Betavalues_MBDIH$Average_D2 <- (Betavalues_MBDIH$Beta_D2_R1 + Betavalues_MBDIH$Beta_D2_R2) / 2
Betavalues_MBDIH$Average_D3 <- (Betavalues_MBDIH$Beta_D3_R1 + Betavalues_MBDIH$Beta_D3_R2) / 2
Betavalues_MBDIH$Average_D4 <- (Betavalues_MBDIH$Beta_D4_R1 + Betavalues_MBDIH$Beta_D4_R2) / 2

# Average for all donors
Betavalues_MBDIH$Average_all <- rowMeans(Betavalues_MBDIH[, c("Average_D1", "Average_D2",
                                                                "Average_D3", "Average_D4")])

# Calculate average beta value per donor — MBD Commercial
Betavalues_MBDCO$Average_D1 <- (Betavalues_MBDCO$Beta_D1_R1 + Betavalues_MBDCO$Beta_D1_R2) / 2
Betavalues_MBDCO$Average_D2 <- (Betavalues_MBDCO$Beta_D2_R1 + Betavalues_MBDCO$Beta_D2_R2) / 2
Betavalues_MBDCO$Average_D3 <- (Betavalues_MBDCO$Beta_D3_R1 + Betavalues_MBDCO$Beta_D3_R2) / 2
Betavalues_MBDCO$Average_D4 <- (Betavalues_MBDCO$Beta_D4_R1 + Betavalues_MBDCO$Beta_D4_R2) / 2

# Average for all donors
Betavalues_MBDCO$Average_all <- rowMeans(Betavalues_MBDCO[, c("Average_D1", "Average_D2",
                                                                "Average_D3", "Average_D4")])

# Create tables with average Epi beta values
Avg_BV_MBDIH <- Betavalues_MBDIH[, c("ProbeIDs", "Average_all")]
Avg_BV_MBDCO <- Betavalues_MBDCO[, c("ProbeIDs", "Average_all")]

# Adding the ATB beta values
ATB_data <- read.csv("2026-02-02_beta_values.csv")

# Calculate average beta value across all ATB donors for all probes
ATB_data$Avg_BV_ATB <- rowMeans(ATB_data[, c("ATB1", "ATB2", "ATB3", "ATB4", "ATB5",
                                               "ATB6", "ATB7", "ATB8", "ATB9", "ATB10")])

# Rename probe column on ATB data frame
names(ATB_data)[1] <- "ProbeIDs"

# Merging beta values for IH
EpiIH_ATB <- merge(Avg_BV_MBDIH,
                   ATB_data[, c("ProbeIDs", "Avg_BV_ATB")],
                   by = "ProbeIDs")

# Merging beta values for CO
EpiCO_ATB <- merge(Avg_BV_MBDCO,
                   ATB_data[, c("ProbeIDs", "Avg_BV_ATB")],
                   by = "ProbeIDs")

# DeltaB for Epi IH
EpiIH_ATB$deltaB_IH <- EpiIH_ATB$Avg_BV_ATB - EpiIH_ATB$Average_all

# DeltaB for Epi CO
EpiCO_ATB$deltaB_CO <- EpiCO_ATB$Avg_BV_ATB - EpiCO_ATB$Average_all

# Calculate average beta values for all Illumina donors
beta$Avg_BV_ill <- rowMeans(beta[, c("D1", "D2", "D3", "D4")])

# Merge average beta values from Illumina and ATB
ill_ATB <- merge(beta[, c("Avg_BV_ill", "ProbeIDs")],
                 ATB_data[, c("ProbeIDs", "Avg_BV_ATB")],
                 by = "ProbeIDs")

# Calculating delta B for ATB vs Illumina
ill_ATB$deltaB_ill <- ill_ATB$Avg_BV_ATB - ill_ATB$Avg_BV_ill

# Saving the data
saveRDS(ill_ATB, file = "ill_ATB_results.rds")

# Inputting the CSV file with top delta B values from study
DeltaB_study <- read.csv("deltaB_study.csv")
colnames(DeltaB_study)[colnames(DeltaB_study) == "Probe_ID"] <- "ProbeIDs"

# Merging delta B values from Illumina, Epi IH, and Epi CO
list_of_dfs <- list(
  ill_ATB[, c("ProbeIDs", "deltaB_ill")],
  EpiIH_ATB[, c("ProbeIDs", "deltaB_IH")],
  EpiCO_ATB[, c("ProbeIDs", "deltaB_CO")],
  DeltaB_study[, c("ProbeIDs", "deltaBeta")])

# Using Reduce to merge the entire list
Top_dmrs_deltaB <- Reduce(function(x, y) merge(x, y, by = "ProbeIDs"), list_of_dfs)

# Check number of rows
nrow(Top_dmrs_deltaB)
# 198 rows
```

## 4.2 PCA plot with all donors and all probe sites

```{r pca-all-donors}
# Merge the data
full_bvalue_list <- list(
  Beta_ill[, c("D1", "D2", "D3", "D4", "ProbeIDs")],
  Betavalues_MBDIH[, c("Beta_D1_R1", "Beta_D1_R2", "Beta_D2_R1", "Beta_D2_R2",
                        "Beta_D3_R1", "Beta_D3_R2", "Beta_D4_R1", "Beta_D4_R2", "ProbeIDs")],
  Betavalues_MBDCO[, c("Beta_D1_R1", "Beta_D1_R2", "Beta_D2_R1", "Beta_D2_R2",
                        "Beta_D3_R1", "Beta_D3_R2", "Beta_D4_R1", "Beta_D4_R2", "ProbeIDs")],
  ATB_data[, c("ATB1", "ATB2", "ATB3", "ATB4", "ATB5", "ATB6", "ATB7", "ATB8", "ATB9", "ATB10",
               "CON1", "CON2", "CON3", "CON4", "CON5", "ProbeIDs")]
)

# Using reduce function to merge data frames
All_Bvalues <- Reduce(function(x, y) merge(x, y, by = "ProbeIDs"), full_bvalue_list)

# Making ProbeIDs the row names
rownames(All_Bvalues)  <- All_Bvalues$ProbeIDs
Matrix_Bvalues <- All_Bvalues[, -which(names(All_Bvalues) == "ProbeIDs")]

# Generating PCA
pca_matrix_BV  <- t(Matrix_Bvalues)
Pca_results_BV <- prcomp(pca_matrix_BV, scale = TRUE)
summary(Pca_results_BV)

# Load library
library(ggplot2)

# Converting PCA results into a data frame for plotting
pca_df_BV <- as.data.frame(Pca_results_BV$x)
pca_df_BV$Sample <- rownames(pca_df_BV)

# Creating groups for legend
pca_df_BV$Group[grepl("D[1-4]", pca_df_BV$Sample)]  <- "Illumina"
pca_df_BV$Group[grepl(".x",     pca_df_BV$Sample)]   <- "Epi_IH"
pca_df_BV$Group[grepl(".y",     pca_df_BV$Sample)]   <- "Epi_CO"
pca_df_BV$Group[grepl("ATB",    pca_df_BV$Sample)]   <- "ATB"
pca_df_BV$Group[grepl("CON",    pca_df_BV$Sample)]   <- "CON"

png("PCA_Plot_Bvalues_All_Samples.png", width = 800, height = 600)

ggplot(pca_df_BV, aes(x = PC1, y = PC2, color = Group)) +
  geom_point(size = 4) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "PCA Plot: Beta Values by Sample Type",
    x     = paste0("PC1 (", round(summary(Pca_results_BV)$importance[2, 1] * 100, 1), "%)"),
    y     = paste0("PC2 (", round(summary(Pca_results_BV)$importance[2, 2] * 100, 1), "%)"),
    color = "Sample Type"
  )

dev.off()
```

## 4.3 Correlation matrix

```{r correlation-matrix}
cd_top_dmr_deltaB <- na.omit(Top_dmrs_deltaB)

# Select only the numeric delta B columns
numeric_cols <- cd_top_dmr_deltaB[, c("deltaB_ill", "deltaB_IH", "deltaB_CO", "deltaBeta")]

# Calculate the Spearman correlation matrix
cor_matrix <- cor(numeric_cols, method = "spearman", use = "pairwise.complete.obs")
print(cor_matrix)

# Making the matrix plot
library(ggplot2)
library(reshape2)

# Melt the correlation matrix
melted_corr <- melt(cor_matrix)

# Create the heatmap
p_melted <- ggplot(data = melted_corr, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low      = "blue",
                       high     = "red",
                       mid      = "white",
                       midpoint = 0,
                       limit    = c(-1, 1),
                       space    = "Lab",
                       name     = "Spearman\nCorrelation") +
  theme_minimal() +
  geom_text(aes(label = round(value, 2))) +
  labs(title = "Methylation Correlation Matrix", x = "", y = "")

# Save the image
ggsave("Fixed_Correlation_matrix.png", plot = p_melted, width = 7, height = 7)
```

**Output — Correlation matrix:**

```
           deltaB_ill  deltaB_IH  deltaB_CO   deltaBeta
deltaB_ill  1.00000000  0.4658454  0.3354381 -0.04468621
deltaB_IH   0.46584545  1.0000000  0.5321246 -0.14873301
deltaB_CO   0.33543814  0.5321246  1.0000000 -0.16798937
deltaBeta  -0.04468621 -0.1487330 -0.1679894  1.00000000
```

## 4.4 Box plots for 7 CpG probe sites

Probes: `cg11985470`, `cg13689801`, `cg20820894`, `cg23240473`, `cg27357902`, `cg07926542`, `cg20250935`

```{r boxplots}
# Rename MBDIH columns
colnames(Betavalues_MBDIH)[10:14] <- paste0(colnames(Betavalues_MBDIH)[10:14], "_IH")

# Rename MBDCO columns
colnames(Betavalues_MBDCO)[10:14] <- paste0(colnames(Betavalues_MBDCO)[10:14], "_CO")

# Create list for mega data frame
megadf_bvalue_list <- list(
  Beta_ill[, c("D1", "D2", "D3", "D4", "ProbeIDs")],
  Betavalues_MBDIH[, c("ProbeIDs", "Average_D1_IH", "Average_D2_IH",
                        "Average_D3_IH", "Average_D4_IH")],
  Betavalues_MBDCO[, c("ProbeIDs", "Average_D1_CO", "Average_D2_CO",
                        "Average_D3_CO", "Average_D4_CO")],
  ATB_data[, c("ATB1", "ATB2", "ATB3", "ATB4", "ATB5", "ATB6", "ATB7", "ATB8",
               "ATB9", "ATB10", "CON1", "CON2", "CON3", "CON4", "CON5", "ProbeIDs")]
)

# Merging the list by ProbeIDs
Megadf_bvalue <- Reduce(function(x, y) merge(x, y, by = "ProbeIDs"), megadf_bvalue_list)

# Load libraries
library(ggplot2)
library(tidyr)
library(dplyr)

# Setup the output folder for box plots
output_folder <- "Boxplot_Beta_DMCs"
if (!dir.exists(output_folder)) dir.create(output_folder)

# Define the probes to plot
target_probes <- c("cg11985470", "cg13689801", "cg20820894",
                   "cg23240473", "cg27357902", "cg07926542", "cg20250935")

# Reshape data from wide to long and assign groups
plot_data_bp <- Megadf_bvalue %>%
  filter(ProbeIDs %in% target_probes) %>%
  pivot_longer(cols = -ProbeIDs, names_to = "Sample", values_to = "Beta_Value") %>%
  mutate(Group = case_when(
    grepl("^D[1-4]$", Sample) ~ "ill",
    grepl("_IH$",     Sample) ~ "Epi_IH",
    grepl("_CO$",     Sample) ~ "Epi_CO",
    grepl("ATB",      Sample) ~ "ATB",
    grepl("CON",      Sample) ~ "Controls_study",
    TRUE                      ~ "Other"
  ))

# Define color palette
group_colors <- c(
  "ATB"            = "#FF0000",
  "ill"            = "#0000FF",
  "Controls_study" = "#0000FF",
  "Epi_IH"         = "#0000FF",
  "Epi_CO"         = "#0000FF"
)

# Run the loop
for (probe in target_probes) {

  current_plot_data <- plot_data_bp %>% filter(ProbeIDs == probe)

  p <- ggplot(current_plot_data, aes(x = Group, y = Beta_Value, fill = Group)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.7) +
    geom_jitter(width = 0.2, alpha = 0.4, color = "black") +
    scale_fill_manual(values = group_colors) +
    theme_minimal() +
    labs(title = paste("Methylation Levels:", probe),
         y     = "Beta Value",
         x     = "Group") +
    theme(axis.text.x    = element_text(angle = 45, hjust = 1),
          legend.position = "none")

  # Save the file
  file_name <- paste0(output_folder, "/", probe, "_colored.png")
  ggsave(filename = file_name, plot = p, width = 8, height = 6, dpi = 300)
}
```
