---
title: "Age Clocks Probes Analysis"
author: "Zoe dell'Osso"
date: "March 2026"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE)
```

---

# 1. Setup & Libraries

```{r libraries}
library(missMethyl)
library(minfi)
library(dplyr)
```

---

# 2. Illumina Data Processing

## Normalization of EPIC Array Data with Noob

```{r noob-normalization}
mSet <- preprocessNoob(rgSet)
```

## Calculate Beta Values from Normalized EPIC Array Data

```{r beta-values}
noob_beta <- getBeta(mSet)
colnames(noob_beta) <- targets$Sample_Name
noob_beta <- as.data.frame(noob_beta)
noob_beta$ProbeIDs <- rownames(noob_beta)
```

## Calculate the Average Beta Value for Each Probe

```{r avg-beta}
noob_beta$Avg_beta <- rowMeans(noob_beta[, c("D1", "D2", "D3", "D4")])
```

---

# 3. Significant Probes for Biological Age

```{r load-ageclock-probes}
ageclock_probes <- read.csv("epigenetic_age_clock_probes.csv")
names(ageclock_probes)[1] <- "ProbeIDs"
```

## Merging

```{r merge-data}
clock_analysis_df <- ageclock_probes %>%
  select(ProbeIDs) %>%
  inner_join(noob_beta[, c("ProbeIDs", "Avg_beta")], by = "ProbeIDs") %>%
  inner_join(Avg_BV_MBDCO[, c("ProbeIDs", "Average_all")], by = "ProbeIDs") %>%
  inner_join(Avg_BV_MBDIH[, c("ProbeIDs", "Average_all")], by = "ProbeIDs")

names(clock_analysis_df)[3] <- "Average_CO"
names(clock_analysis_df)[4] <- "Average_IH"
```

---

# 4. Confusion Matrix

```{r confusion-matrix-libraries}
# -- loading libraries --
library(dplyr)
library(tidyr)
library(caret)
library(ggplot2)
library(caret)
library(reshape2)
```

## In House

```{r confusion-matrix}
# creating the 10 different classes with 0.1 intervals between each
my_breaks3 <- c(0, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1.0)
my_labels3 <- c("Class 1", "Class 2", "Class 3", "Class 4", "Class 5",
                 "Class 6", "Class 7", "Class 8", "Class 9", "Class 10")

# classifying the beta values into the 10 classes
matrixdf_c10_clock_analysis_df_IH <- clock_analysis_df %>% mutate(Epi_Class = cut(Average_IH, breaks = my_breaks3, labels = my_labels3, include.lowest = TRUE), Illumina_Class = cut(Avg_beta, breaks = my_breaks3, labels = my_labels3, include.lowest = TRUE))

# making confusion matrix
conf_matrix_class10_clock_analysis_df_IH <- confusionMatrix(
  data      = matrixdf_c10_clock_analysis_df_IH$Epi_Class,
  reference = matrixdf_c10_clock_analysis_df_IH$Illumina_Class)

# converting the confusion matrix table into a df
plotdata_cm_class10_clock_analysis_df_IH <- as.data.frame(conf_matrix_class10_clock_analysis_df_IH$table)

# plotting the confusion matrix
plot_cm_class10_clock_analysis_df_IH <- ggplot(plotdata_cm_class10_clock_analysis_df_IH, aes(x = Prediction, y = Reference, fill = Freq)) + geom_tile(color = "white", lwd = 1) + geom_text(aes(label = Freq), color = "black", size = 3.5) + scale_fill_gradient(low = "#ebf3fb", high = "#3182bd") +  labs( x = "Epi IH", y = "Illumina", title = "Confusion Matrix: Epi IH vs Illumina", fill = "Count" ) + theme_minimal() + theme( axis.text = element_text (angle = 45, vjust = 1, hjust = 1), title = element_text(size = 14, face = "bold"), aspect.ratio = 1 )

# saveing as a PNG
ggsave("confusion_matrix_c10_clock_analysis_df_IH.png", plot = plot_cm_class10_clock_analysis_df_IH, width = 7, height = 6, dpi = 300)
```

## Commercial

```{r confusion-matrix-CO}
# creating the 10 different classes with 0.1 intervals between each
my_breaks3 <- c(0, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1.0)
my_labels3 <- c("Class 1", "Class 2", "Class 3", "Class 4", "Class 5",
                 "Class 6", "Class 7", "Class 8", "Class 9", "Class 10")

# classifying the beta values into the 10 classes
matrixdf_c10_clock_analysis_df_CO <- clock_analysis_df %>% mutate(Epi_Class = cut(Average_CO, breaks = my_breaks3, labels = my_labels3, include.lowest = TRUE), Illumina_Class = cut(Avg_beta, breaks = my_breaks3, labels = my_labels3, include.lowest = TRUE))

# making confusion matrix
conf_matrix_class10_clock_analysis_df_CO <- confusionMatrix(
  data      = matrixdf_c10_clock_analysis_df_CO$Epi_Class,
  reference = matrixdf_c10_clock_analysis_df_CO$Illumina_Class)

# converting the confusion matrix table into a df
plotdata_cm_class10_clock_analysis_df_CO <- as.data.frame(conf_matrix_class10_clock_analysis_df_CO$table)

# plotting the confusion matrix
plot_cm_class10_clock_analysis_df_CO <- ggplot(plotdata_cm_class10_clock_analysis_df_CO, aes(x = Prediction, y = Reference, fill = Freq)) + geom_tile(color = "white", lwd = 1) + geom_text(aes(label = Freq), color = "black", size = 3.5) + scale_fill_gradient(low = "#ebf3fb", high = "#3182bd") +  labs( x = "Epi CO", y = "Illumina", title = "Confusion Matrix: Epi CO vs Illumina", fill = "Count" ) + theme_minimal() + theme( axis.text = element_text (angle = 45, vjust = 1, hjust = 1), title = element_text(size = 14, face = "bold"), aspect.ratio = 1 )

# saveing as a PNG
ggsave("confusion_matrix_c10_clock_analysis_df_CO.png", plot = plot_cm_class10_clock_analysis_df_CO, width = 7, height = 6, dpi = 300)
```

---

# 5. GO Enrichment Analysis on Misclassified Probes

```{r go-libraries}
library(dplyr)
library(missMethyl)
```

## In House

### Identify Misclassified Probes

```{r misclassified-probes}
misclassified_probes_df <- matrixdf_c10_clock_analysis_df_IH %>%
  filter(Epi_Class != Illumina_Class)

misclassified_ids <- misclassified_probes_df$ProbeIDs
length(misclassified_ids)
```

**Output:** `1267`

### Run GO Analysis Based on 17K Probes Associated with Age Clock

```{r go-analysis}
all_probes <- clock_analysis_df$ProbeIDs

gst_errors <- gometh(
  sig.cpg    = misclassified_ids,
  all.cpg    = matrixdf_c10_clock_analysis_df_IH$ProbeIDs,
  collection = "GO",
  array.type = "EPIC")

top_error_terms <- topGSA(gst_errors, n = 10)
print(top_error_terms)
```

**Output — Top 10 GO Terms:**

```
           ONTOLOGY                                    TERM   N  DE        P.DE         FDR
GO:0009617       BP                   response to bacterium  64  56 0.001861123           1
GO:0005102       MF              signaling receptor binding 142 116 0.003127581           1
GO:0001508       BP                        action potential  17  17 0.004110199           1
GO:0007218       BP          neuropeptide signaling pathway  16  16 0.004789182           1
GO:0001906       BP                            cell killing  16  16 0.005572132           1
GO:0042742       BP           defense response to bacterium  22  21 0.006014938           1
GO:0040011       BP                              locomotion 131 106 0.008808557           1
GO:0072659       BP protein localization to plasma membrane  22  21 0.009284409           1
GO:0006936       BP                      muscle contraction  42  37 0.009650699           1
GO:1901981       MF  phosphatidylinositol phosphate binding  20  19 0.011844948           1
```

### Generating GO Enrichment Graph

```{r go-enrichment-plot}
library(ggplot2)
library(dplyr)

gorich_plot_data <- topGSA(gst_errors, n = 15) %>%
  mutate(Term = factor(TERM, levels = rev(TERM)),
         GeneRatio = DE / N)

goenrich_plot_top15 <- ggplot(gorich_plot_data, aes(x = GeneRatio, y = Term)) + geom_point(aes(size = DE, color = P.DE)) + scale_color_gradient(low = "red", high = "blue") + theme_minimal() + labs(title = "GO Enrichment: Misclassified Probes for Epi IH", subtitle = "Top 15 Biological Terms by Raw P-value", x = "Gene Ratio (DE / N)", y = "Biological Process / Function", size = "Probe Count", color = "P-value") 

ggsave("goenrich_plot_top15.png",
       plot   = goenrich_plot_top15,
       width  = 10,
       height = 8,
       dpi    = 300)
```

## Commercial

### Identify Misclassified Probes

```{r misclassified-probes-CO}
misclassified_probes_df_CO <- matrixdf_c10_clock_analysis_df_CO %>%
  filter(Epi_Class != Illumina_Class)

misclassified_ids_CO <- misclassified_probes_df_CO$ProbeIDs
length(misclassified_ids_CO)
```

**Output:** `1230`

### Run GO Analysis Based on 17K Probes Associated with Age Clock

```{r go-analysis-CO}
all_probes_CO <- clock_analysis_df$ProbeIDs

gst_errors_CO <- gometh(
  sig.cpg    = misclassified_ids_CO,
  all.cpg    = matrixdf_c10_clock_analysis_df_CO$ProbeIDs,
  collection = "GO",
  array.type = "EPIC")

top_error_terms_CO <- topGSA(gst_errors_CO, n = 10)
print(top_error_terms_CO)
```

**Output — Top 10 GO Terms:**

```
           ONTOLOGY                                                TERM   N  DE
GO:0048762       BP                    mesenchymal cell differentiation  29  27
GO:0098771       BP                           inorganic ion homeostasis  52  45
GO:0016477       BP                                      cell migration 143 113
GO:0050801       BP                          monoatomic ion homeostasis  54  46
GO:0055080       BP                       monoatomic cation homeostasis  54  46
GO:0040011       BP                                          locomotion 131 103
GO:0055074       BP                             calcium ion homeostasis  34  30
GO:0031349       BP             positive regulation of defense response  39  34
GO:0031345       BP negative regulation of cell projection organization  19  18
GO:0001837       BP                epithelial to mesenchymal transition  19  18
                  P.DE FDR
GO:0048762 0.001974829   1
GO:0098771 0.002641135   1
GO:0016477 0.003744580   1
GO:0050801 0.005113007   1
GO:0055080 0.005113007   1
GO:0040011 0.006680243   1
GO:0055074 0.006901164   1
GO:0031349 0.007802787   1
GO:0031345 0.008147444   1
GO:0001837 0.008664810   1
```

### Generating GO Enrichment Graph

```{r go-enrichment-plot-CO}
library(ggplot2)
library(dplyr)

gorich_plot_data_CO <- topGSA(gst_errors_CO, n = 15) %>%
  mutate(Term = factor(TERM, levels = rev(TERM)),
         GeneRatio = DE / N)

goenrich_plot_top15_CO <- ggplot(gorich_plot_data_CO, aes(x = GeneRatio, y = Term)) + geom_point(aes(size = DE, color = P.DE)) + scale_color_gradient(low = "red", high = "blue") + theme_minimal() + labs(title = "GO Enrichment: Misclassified Probes for Epi CO", subtitle = "Top 15 Biological Terms by Raw P-value", x = "Gene Ratio (DE / N)", y = "Biological Process / Function", size = "Probe Count", color = "P-value") 

ggsave("goenrich_plot_top15_CO.png",
       plot   = goenrich_plot_top15_CO,
       width  = 10,
       height = 8,
       dpi    = 300)
```
---

# 6. Distribution Analysis

## Assessing if Normally Distributed

```{r distribution-libraries}
library(ggplot2)
library(tidyverse)
```

```{r distribution-plot}
plot_long <- clock_analysis_df %>%
  pivot_longer(cols = c(Avg_beta, Average_CO, Average_IH),
               names_to  = "Method",
               values_to = "Beta_Value")

clockprobes_distribution <- ggplot(plot_long,
                                    aes(x = Beta_Value, color = Method)) +
  geom_line(stat = "density")

# saving the histogram as a PNG
ggsave("clockprobes_distribution.png",
       plot   = clockprobes_distribution,
       width  = 10,
       height = 6,
       dpi    = 300)
```

---

# 7. Linear Relationship Assessment

As the data is not normally distributed, we assess whether a linear relationship is present.

## Commercial

```{r scatterplot-commercial}
library(ggplot2)

clockprobes_scatterplotCO <- ggplot(clock_analysis_df,
                                     aes(x = Avg_beta, y = Average_CO)) +
  geom_point(alpha = 0.2) +
  xlim(0, 1) +
  ylim(0, 1) +
  labs(
    title    = "Beta Value Comparison: MBD Commercial vs. Array",
    subtitle = "Red dashed line indicates perfect agreement (y = x)",
    x        = "Array (Avg_beta)",
    y        = "MBD Commercial (Average_CO)"
  ) +
  theme_minimal()

ggsave("clockprobes_scatterplotCO.png",
       plot   = clockprobes_scatterplotCO,
       width  = 10,
       height = 6,
       dpi    = 300)
```

## In House

```{r scatterplot-inhouse}
clockprobes_scatterplotIH <- ggplot(clock_analysis_df,
                                     aes(x = Avg_beta, y = Average_IH)) +
  geom_point(alpha = 0.1) +
  theme_minimal() +
  labs(
    title    = "Beta Value Comparison: MBD In House vs. Array",
    subtitle = "Red dashed line indicates perfect agreement (y = x)",
    x        = "Array (Avg_beta)",
    y        = "MBD Commercial (Average_IH)"
  ) +
  xlim(0, 1) +
  ylim(0, 1)

ggsave("clockprobes_scatterplotIH.png",
       plot   = clockprobes_scatterplotIH,
       width  = 10,
       height = 6,
       dpi    = 300)
```

---

# 8. Correlation Matrix Using Spearman's Rank Correlation

```{r spearman-correlation}
cor_matrix <- cor(cor_data, method = "spearman", use = "pairwise.complete.obs")
head(cor_matrix)
```

**Output:**

```
                   Array MBD Commercial MBD In House
Array          1.0000000      0.7536055    0.6574051
MBD Commercial 0.7536055      1.0000000    0.6988752
MBD In House   0.6574051      0.6988752    1.0000000
```

```{r correlation-heatmap}
library(ggplot2)
library(reshape2)

# Melt the correlation matrix
melted_corr <- melt(cor_matrix)

# Create the heatmap
corr_heatmap <- ggplot(melted_corr, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0) +
  theme_minimal() +
  labs(title = "Spearman Correlation Heatmap",
       x = "", y = "", fill = "Correlation")

ggsave("clockprobes_spearman_matrix.png",
       plot   = corr_heatmap,
       width  = 7,
       height = 6,
       dpi    = 300)
```

---

